---
title: "University of Oregon Running Game"
author: "Quinn MacLean"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load data in future 
library(kableExtra)
library(ggridges)
library(dplyr)
library(RColorBrewer)
library(ggpmisc)
library(cfbscrapR)

pbp<-readRDS(file = "cfb_pbp.rds")
play_df<-readRDS(file="oregonplay_df.rds")



#2014 - 2020
o_rush_pbp<-pbp %>%
  select(game_id,year,pos_team,rush,half,period,down,EPA,yards_gained,rusher_player_name,success,play_type,venue,yard_line) %>%
  filter(pos_team == "Oregon",play_type != "2pt Conversion",play_type != "Extra Point Missed")




o_rush_pbp_prev<-play_df %>%
  select(game_id,season,pos_team,rush,half,period,down,ppa,yards_gained,rush_player,success,play_type,venue,yard_line) %>%
  filter(pos_team == "Oregon",play_type != "2pt Conversion",play_type != "	Extra Point Missed")

names(o_rush_pbp_prev)[2]<-"year"
names(o_rush_pbp_prev)[8]<-"EPA"
names(o_rush_pbp_prev)[10]<-"rusher_player_name"
names(o_rush_pbp_prev)[13]<-"venue"

o_rush_pbp<-rbind(o_rush_pbp,o_rush_pbp_prev)

HC_year<-c(2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2020)
HC_name<-c("Chip Kelly","Chip Kelly","Chip Kelly",
           "Mark Helfrich","Mark Helfrich","Mark Helfrich","Mark Helfrich","Willie Taggart","Mario Cristobal","Mario Cristobal","Mario Cristobal")

HC<-cbind(HC_year,HC_name)

OC_name<-c("Mark Helfrich","Mark Helfrich","Mark Helfrich","Scott Frost","Scott Frost","Scott Frost","Matt Lubick","Mario Cristobal/Marcus Arroyo","Marcus Arroyo",
           "Marcus Arroyo","Joe Moorhead")

HC<-cbind(HC,OC_name)

#unique(pbp$offense_conference)
### PAC 12 Avgs 2014 -2020 #####
o_rush_pbp_pc12<-pbp %>%
  select(game_id,year,pos_team,rush,half,period,down,EPA,yards_gained,rusher_player_name,success,play_type,venue,yard_line,offense_conference) %>%
  filter(offense_conference == "Pac-12",play_type != "2pt Conversion",play_type != "Extra Point Missed")

#2010 - 2013
#unique(play_df$offense_conference)
o_rush_pbp_prev_pc12<-play_df %>%
  select(game_id,season,pos_team,rush,half,period,down,ppa,yards_gained,rush_player,success,play_type,venue,yard_line,offense_conference) %>%
  filter(offense_conference %in% c('Pac-12','Pac-10'),play_type != "2pt Conversion",play_type != "	Extra Point Missed")

names(o_rush_pbp_prev_pc12)[2]<-"year"
names(o_rush_pbp_prev_pc12)[8]<-"EPA"
names(o_rush_pbp_prev_pc12)[10]<-"rusher_player_name"
names(o_rush_pbp_prev_pc12)[13]<-"venue"

o_rush_pbp_pc12<-rbind(o_rush_pbp_pc12,o_rush_pbp_prev_pc12)

```

# Introduction

The purpose of this analysis is to analyze University of Oregon football's running game over time. We aim to determine the following: \
1. How's the rushing offense been trending?  \
2. Who's the most effective & efficient runner? \
3. Identify other interesting tidbits \
\

The chart below shows rushing plays during the 2010-2020 timeframe. In the timeframe, Oregon competed in 2 national championships. 2010, 2014 in the following charts show those championship runs. EPA is the expected or predicted points per attempt (2010-2013, we used PPA, 2014-2020 we used EPA), this means how much points could have results from any rush attempt. A successful run is more than more than 50% of distance to 1st down rushing yards on 1st down (i.e. First down with 10 yards to go, a rush of over 5 yards is successful run), more than 70% on 2nd down and distance to go, or if the 3rd or 4th down run resulted in a first down. The success rate would yield the amount of a successful runs to attempts. YPC is the average yards gained during rush attempt, and plays is the total number of rushing attempts on a given season. It should be noted that we filtered out any 2-point conversion attempts as not part of the analysis. 

# Team Analysis
We can see that in the Championship runs the running game was very identical with the exception that 2014 had a higher mean_EPA meaning much more explosive plays per attempt. Keep that in mind when we analyze players performance. The Chip Kelly years was much more rush heavy (600+ rush attempts). You can see the Mark Helfrich/Scott Frost combo was actually the most explosive in the 1st year (2013) and most efficient at 53% success rate and a 6.6 yards gained average in less than 100 rush attempts in the previous season. You could see the offense started to shy away from the rush game. We can see it's less successful but the EPA went up meaning it was a much more explosive rushing attack. Largely the 2014 campaign relied on explosive runs to make it the National championship. The removal of either Mark Helfrich or Scott Frost involved in the offense, we can say that Oregon has never been the same. The yards per attempt in 2020 reach 2014 levels but the expected points per attempt isn't good, which probably means the offense lacked any explosiveness. Something to keep an eye on in 2020. 


```{r a1,echo=FALSE,message=FALSE,warning=FALSE}

a1<-o_rush_pbp %>%
  filter(pos_team == "Oregon",rush==1,down<=4) %>%
  group_by(pos_team,year) %>%
  summarize(mean_epa = mean(EPA,na.rm = TRUE),
            success_rate = mean(success),
            ypc=mean(yards_gained),
            plays=n()) %>%
  arrange(-desc(year)) %>%
  filter(plays>50)

a1[,3:5]<-round(a1[,3:5],2)
HC<-as.data.frame(HC)
names(HC)[1]<-"year"
HC$year<-as.numeric(as.character(HC$year))

a1<-left_join(HC,a1,by=c("year" = "year"))

pc12avg<-o_rush_pbp_pc12 %>%
  filter(rush==1,down<=4,pos_team != "Oregon") %>%
  group_by(year) %>%
  summarize(mean_epa = mean(EPA,na.rm = TRUE),
            success_rate = mean(success),
            ypc=mean(yards_gained),
            plays=n()) %>%
  arrange(-desc(year)) %>%
  filter(plays>50)

pc12avg[,2:4]<-round(pc12avg[,2:4],2)

a1<-left_join(pc12avg,a1,by=c("year"="year"))

names(a1)[2]<-"Pac12_mean_EPA"
names(a1)[3]<-"Pac12_success_rate"
names(a1)[4]<-"Pac12_YPC"
names(a1)[5]<-"Pac12_Plays"
names(a1)[9]<-"mean_epa"
names(a1)[10]<-"success_rate"
names(a1)[11]<-"ypc"
names(a1)[12]<-"plays"

a1<- a1 %>%
  select(year,pos_team,HC_name,OC_name,mean_epa,Pac12_mean_EPA,success_rate,Pac12_success_rate,ypc,Pac12_YPC,plays,Pac12_Plays)

names(a1)[5]<-"Oregon_EPA"
names(a1)[6]<-"PAC12_EPA"
names(a1)[7]<-"Oregon"
names(a1)[8]<-"PAC12"
names(a1)[9]<-"Oregon YPC"
names(a1)[10]<-"PAC12 YPC"
names(a1)[11]<-"Oregon Plays"
names(a1)[12]<-"PAC12 Plays"

a1 %>%
kable(caption = "Oregon Running Game vs rest of PAC over time 2010-2020") %>%
kable_styling(font_size = 10,bootstrap_options = c("condensed")) %>%
   add_header_above(c(" " = 4, "EPA" = 2, "Success Rate" = 2, "YPC" = 2,"Plays"=2))


## few things going on here; Herbert was a very good runner
#with 7.6 yard per play. Tyler Shough is actually a decent 
# runner but only


```
#### Comparison to Passing Game Effectivness
We can see the rush effectiveness & yards gained less steeper of a line than the passing attack meaning much more explosiveness came from the rushing attack over the history. 
```{r a1a,echo=FALSE,message=FALSE,warning=FALSE}

o<-o_rush_pbp %>%
  filter(pos_team == "Oregon",down<=4) %>%
  mutate(play_type = case_when(
    play_type == "Rushing Touchdown" ~ "Rush",
    play_type %in% c("Pass Incompletion","Pass Completion","Passing Touchdown","Pass Reception") ~ "Pass",
    TRUE ~ as.character(play_type)
  ))

my.formula<- y~ x
o %>%
  filter(play_type %in% c('Rush','Pass')) %>%
  ggplot(aes(y=yards_gained,x=EPA,color=year)) +
           geom_point() +
  facet_wrap(~play_type) +
  theme(legend.position = "none",
        legend.text = element_text(color="black",size=7)) +
  geom_smooth(method="lm",se=FALSE,color="green",formula=my.formula) +
   stat_poly_eq(formula = my.formula, 
                aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")), 
                parse = TRUE) +
  ggtitle("Oregon Pass vs. Rush")
  

```

##### Rushing Attack by Stadium
The Ducks are at their best in terms of rushing attack when they head up the I-5 to Reser Stadium where they average nearly 7 yards gained per attempt and have a successful run nearly 50% of the time. The 0.2 EPA means they have a lot of breakout plays at Reser Stadium. The Ducks struggles in running game when traveling to Memorial Stadium, home of Cal Bears and Sun Devil Stadium. The game at Memorial Stadium, home of U of Nebraska was heartbreaking despite Oregon's most success game running the football. Lastly, the AT&T curse is real for the Ducks rushing attack as there is negative Expected points per attempt. 
```{r a1b,echo=FALSE,message=FALSE,warning=FALSE}

b<-subset(o_rush_pbp,o_rush_pbp$venue == "")

a2<-o_rush_pbp %>%
  filter(pos_team == "Oregon",rush==1,down<=4,complete.cases(venue),venue!="") %>%
  group_by(venue) %>%
  summarize(mean_epa = mean(EPA,na.rm = TRUE),
            success_rate = mean(success),
            ypc=mean(yards_gained),
            plays=n()) %>%
  arrange(desc(plays)) %>%
  filter(plays>50)

a2[,2:4]<-round(a2[,2:4],2)

a2 %>%
kable(caption = "Oregon Running Game by Venue") %>%
kable_styling(font_size = 10,bootstrap_options = c("condensed"))
  

```
#### Rushing Effectiveness by Down
Data on 1st & 2nd down is much more predictably given the R^2 is above 0.5. That being said Oregon was much more efficient running on 1st down then on 2nd down with the a higher amount of yards gained per EPA. 
```{r o,echo=FALSE,message=FALSE,warning=FALSE}

o_rush<-o_rush_pbp %>%
  filter(pos_team == "Oregon",down<=4,rush==1,down>=1)

my.formula<- y~ x
o_rush %>%
  ggplot(aes(y=yards_gained,x=EPA,color=year)) +
           geom_point() +
  facet_wrap(~down) +
  theme(legend.position = "none",
        legend.text = element_text(color="black",size=7)) +
  geom_smooth(method="lm",se=FALSE,color="green",formula=my.formula) +
   stat_poly_eq(formula = my.formula, 
                aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")), 
                parse = TRUE) +
  ggtitle("Oregon Rush Effectiveness by Down")


```

#### Oregon Rushing on 1st Down
Oregon rushing on 1st down hasn't been much of a thing since Mario Cristobal took over in 2018. It was very much a strategy of Willie Taggart and although not always successful it yield some high explosive plays. 2014 was the 2nd most explosive on 1st down but 2011 was likely to best year overall in ranking of stats overall. 
```{r o1,echo=FALSE,message=FALSE,warning=FALSE}


a3<-o_rush_pbp %>%
  filter(pos_team == "Oregon",rush==1,down==1) %>%
  group_by(year) %>%
  summarize(mean_epa = mean(EPA,na.rm = TRUE),
            success_rate = mean(success),
            ypc=mean(yards_gained),
            plays=n()) %>%
  arrange(-desc(year)) %>%
  filter(plays>50)

a3[,2:4]<-round(a3[,2:4],2)


a4<-o_rush_pbp_pc12 %>%
  filter(rush==1,down==1,pos_team != "Oregon") %>%
  group_by(year) %>%
  summarize(mean_epa = mean(EPA,na.rm = TRUE),
            success_rate = mean(success),
            ypc=mean(yards_gained),
            plays=n()) %>%
  arrange(-desc(year)) %>%
  filter(plays>50)

a4[,2:4]<-round(a4[,2:4],2)

a3<-left_join(a4,a3,by=c("year"="year"))

names(a3)[2]<-"PAC12_mean_EPA"
names(a3)[3]<-"PAC12_success_rate"
names(a3)[4]<-"PAC12_ypc"
names(a3)[5]<-"PAC12_plays"
names(a3)[6]<-"mean_EPA"
names(a3)[7]<-"success_rate"
names(a3)[8]<-"ypc"
names(a3)[9]<-"plays"

a3<-a3 %>%
  select(year,mean_EPA,PAC12_mean_EPA,success_rate,PAC12_success_rate,ypc,PAC12_ypc,plays,PAC12_plays)

names(a3)[2]<-"Oregon EPA"
names(a3)[3]<-"PAC 12 EPA"
names(a3)[4]<-"Oregon"
names(a3)[5]<-"PAC 12"
names(a3)[6]<-"Oregon YPC"
names(a3)[7]<-"PAC 12 YPC"
names(a3)[8]<-"Oregon Plays"
names(a3)[9]<-"PAC 12 Plays"


a3 %>%
kable(caption = "Oregon vs PAC Rushing on 1st down") %>%
kable_styling() %>%
kable_styling(font_size = 10,bootstrap_options = c("condensed")) %>%
   add_header_above(c(" " = 1, "EPA" = 2, "Success Rate" = 2, "YPC" = 2,"Plays"=2))



```


# Individual Career Performances
The next session we start to take a look at individual career performances at UO, which is combo between QB/RB and hybrid RB/WR.
First thing you notice is the explosiveness of Marcus Mariota and his high success rate. Makes sense given he won the Heisman Trophy in 2014. The fact that per Rushing attempt in the entirety of his career he nearly put a half a point on the scoreboard is as the expected result of his rushing attempt. Per 5 attempts he was the equivalent to a FG kicker. To also add this to perspect, Aaron Rodger's 2020 campaign was EPA per play in modern NFL at 0.35 so 0.91 on the rushing attack shows how prolific he was. Plus he averaged nearly 9 yards gained per attempt on over 229 rush attempts. WILD. \

#### Career High EPA Runners
We can see the QBs are at the top of the efficient runners, which makes sense given the historical read player option variation of the offense. Tony Brooks-James if one of the more efficient runner despite less attempts than Marcus Mariota. Not surprising, De'Anthony Thomas had a relatively high Mean EPA despite low touches, he was the definition of explosiveness in Oregon history and arguably all first team kick return/punt return. LaMichael's data is only based on two seasons of data but arguably his best two seasons in program history. His partner in crime, Kenjon Barner is arguably the more explosive and effective runner of the duo. One note of optimism is Tyler Shough, he was an effective runner off less plays BUT at his rate per season he's likely approaching Marcus Mariota rushing attempts if he was a 4 year starter. 
```{r p1,echo=FALSE,message=FALSE,warning=FALSE}

oregon<-o_rush_pbp %>%
  filter(pos_team == "Oregon",rush==1,down<=4,complete.cases(EPA)) %>%
  group_by(rusher_player_name) %>%
  summarize(mean_epa = mean(EPA),
            success_rate = mean(success),
            ypc=mean(yards_gained),
            plays=n(),
            `seasons of data` = n_distinct(year)) %>%
  arrange(desc(mean_epa)) %>%
  filter(plays>50) %>%
  head(n=15L)
  

names(oregon)[1]<-"Rushers"
oregon[,2:4]<-round(oregon[,2:4],2)

oregon %>%
kable(caption = "Top Oregon Historical EPA Runners 2010-2020") %>%
  kable_styling(font_size = 10,bootstrap_options = c("condensed")) %>%
  footnote(general = "Filtered for Top 15 career")


or<-o_rush_pbp %>%
  filter(pos_team == "Oregon",rush==1,down<=4,complete.cases(yards_gained),
         complete.cases(rusher_player_name)) %>%
  group_by(rusher_player_name)  %>%
  filter(n()>50)




```

#### PAC 12 Workhorses
Royce Freeman and Myles Gaskin had a similar career in the PAC 12 with Royce Freem gaining the edge. Ronald Jobes & Christian McCaffrey were the most explosive in their career. Zach Moss could be argued to be the most consistent of the bunch given a high success rate. 
```{r p1b,echo=FALSE,message=FALSE,warning=FALSE}


p12<-o_rush_pbp_pc12 %>%
  filter(offense_conference %in% c('Pac-12','Pac-10'),rush==1,down<=4,complete.cases(EPA)) %>%
  group_by(rusher_player_name,pos_team) %>%
  summarize(mean_epa = mean(EPA),
            success_rate = mean(success),
            ypc=mean(yards_gained),
            plays=n(),
            `seasons of data` = n_distinct(year)) %>%
  arrange(desc(plays)) %>%
  filter(plays>100) %>%
  head(n=15L)

names(p12)[1]<-"Rushers"
p12[,3:5]<-round(p12[,3:5],2)

p12 %>%
kable(caption = "Pac12 Workhorses 2010-2020") %>%
  kable_styling(font_size = 10,bootstrap_options = c("condensed")) %>%
  footnote(general = "Top 15 by Career Rushes")




```

We can see that PAC 12 QBs dominate the EPA given their overall efficiency. It may be something to say that Arizona's running game was may explosive with a running QB than their primary back. Not surprising Marcus Mariota was one of the most explosive QBs in the past decade in terms of rushing attack. Surprising that Max Borghi (the sole Running back below) was the most explosive runner given he was in an air raid offense. 
```{r p1c,echo=FALSE,message=FALSE,warning=FALSE}

p12<-o_rush_pbp_pc12 %>%
  filter(offense_conference %in% c('Pac-12','Pac-10'),rush==1,down<=4,complete.cases(EPA)) %>%
  group_by(rusher_player_name,pos_team) %>%
  summarize(mean_epa = mean(EPA),
            success_rate = mean(success),
            ypc=mean(yards_gained),
            plays=n(),
            `seasons of data` = n_distinct(year)) %>%
  arrange(desc(mean_epa)) %>%
  filter(plays>100) %>%
  head(n=15L)

names(p12)[1]<-"Rushers"
p12[,3:5]<-round(p12[,3:5],2)

p12 %>%
kable(caption = "Top Pac12 EPA 2010-2020") %>%
  kable_styling(font_size = 10,bootstrap_options = c("condensed")) %>%
  footnote(general = "Top 15 EPA or explosive runners")
  

```

#### PAC12 EPA v Success Rate 
This next section charts their explosiveness (EPA) around their success rate. We can see the clear separation of Marcus Mariota. In fact all Oregon runningbacks are colored in green below. One obvious question is will Oregon ever get a gamechanging runner like Marcus Mariota? Justin Herbert's 2019 Rose Bowl performance (3 Rushing TDs) left a big what-if for a lot of Oregon Duck fans. 
```{r p1d,echo=FALSE,message=FALSE,warning=FALSE}

library(ggrepel)

pc12 <- o_rush_pbp_pc12 %>% 
	mutate(
		name = rusher_player_name,
		rusher = rusher_player_name,
		play = 1
	) %>%
	group_by(name, pos_team) %>%
	summarize (
		n_rush = sum(rush),
		n_plays = sum(play),
		epa_per_play = mean(EPA,na.rm = TRUE),
		success_per_play =mean(success),
		rush_yds = sum(yards_gained),
		) %>%
	filter(n_plays>=100,complete.cases(name))


pc12 %>%
  ggplot(aes(x = success_per_play, y = epa_per_play)) +
  geom_hline(yintercept = mean(pc12$epa_per_play), color = "red", linetype = "dashed") +
  geom_vline(xintercept =  mean(pc12$success_per_play), color = "red", linetype = "dashed") +
  geom_point(color = ifelse(pc12$pos_team == "Oregon", "green", "black"), cex=pc12$n_plays/60, alpha=1/16) +
  geom_text_repel(aes(label=name),
      force=1, point.padding=0,
      segment.size=0.1) +
  labs(x = "Success rate",
       y = "EPA per play",
       caption = "Data from cfbscrapR",
       title = "PAC12 success rate and EPA/play",
       subtitle = "2010-2020") +
  theme_bw() +
  theme(axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        plot.title = element_text(size = 16, hjust = 0.5),
        plot.subtitle = element_text(size = 14, hjust = 0.5),
        plot.caption = element_text(size = 12))




```

#### PAC12 Career Rush Yds vs. EPA
We can see the overall consistency of Royce Freeman and Myles Gaskin, which made for a great rivalry the years they squared off. Again, notice Marcus Mariota's explosiveness compared to other players on the chart below. 
```{r p1e,echo=FALSE,message=FALSE,warning=FALSE}

pc12 %>%
  ggplot(aes(x = rush_yds, y = epa_per_play)) +
  geom_hline(yintercept = mean(pc12$epa_per_play), color = "red", linetype = "dashed") +
  geom_vline(xintercept =  mean(pc12$rush_yds), color = "red", linetype = "dashed") +
  geom_point(color = ifelse(pc12$pos_team == "Oregon", "green", "black"), cex=pc12$n_plays/60, alpha=1/16) +
  geom_text_repel(aes(label=name),
      force=1, point.padding=0,
      segment.size=0.1) +
  labs(x = "Rush Yards",
       y = "EPA per play",
       caption = "Data from cfbscrapR",
       title = "PAC12 Career Rush Yds and EPA/play",
       subtitle = "2010-2020") +
  theme_bw() +
  theme(axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        plot.title = element_text(size = 16, hjust = 0.5),
        plot.subtitle = element_text(size = 14, hjust = 0.5),
        plot.caption = element_text(size = 12))


```


#### Yards Gained Distribution
The charts show average distribution comparison of yards gained per Rushers. One interesting observation, which was confirmed above the clear higher than average attempt of Marcus Mariota. His distribution of runs featured more break away, explosive gains than others. Not surprisingly. Justin Herbert and De'Anthony Thomas averaged had a similar distribution of yards gained, which is interested given their physiques are a stark contrast from each other. 
```{r pressure,echo=FALSE,message=FALSE,warning=FALSE}

or<-o_rush_pbp %>%
  filter(pos_team == "Oregon",rush==1,down<=4,complete.cases(yards_gained),
         complete.cases(rusher_player_name)) %>%
  group_by(rusher_player_name)  %>%
  mutate(yds_cum = cumsum(yards_gained)) %>%
  filter(n()>100) %>%
  arrange(desc(yds_cum))

nb.cols<-15
mycolors<-colorRampPalette(brewer.pal("Blues","Greens"))(nb.cols)

ggplot(or,aes(x = yards_gained, y = rusher_player_name, fill=rusher_player_name)) +
  geom_density_ridges(quantile_lines=TRUE,
                      quantile_fun=function(x,...)mean(x)) +
  theme(legend.position = "none") +
  geom_text(data=or %>%
              group_by(rusher_player_name) %>%
              summarise(yards_gained=median(yards_gained)),
aes(label=sprintf("%1.1f",yards_gained),
position=position_nudge(y=-0.1), color="red",size=3.5)) %>%
  scale_fill_manual(values = mycolors) +
  xlab("Yards Gained") +
  ylab("Distribution of Oregon Runs") +
labs(title = "Oregon Efficiency Runners",
              subtitle = "2014-2020; >50 Rushes",
              caption = "Data source: CFBScraper")
  
  
```

# Individual Oregon Season Performances
Mariota's 2014 season was clearly superb and big reason for why he won the Heisman. The fact he makes up the top 3 explosive running season is purely phenomenal. 2011 Rose Bowl was largely driven a high efficient attack of many runners. CJ Verdell's 2020 was not good whatsover as it was a stark difference from his 2019 season where his draft stock was rising. His decision to return, we'd hopefully like to see his 2019 season and not his 2020 season (hampered by injuries)

```{r a2a,echo=FALSE,message=FALSE,warning=FALSE}

oregon_year<-o_rush_pbp %>%
  filter(pos_team == "Oregon",rush==1,down<=4,complete.cases(EPA)) %>%
  group_by(rusher_player_name,year) %>%
  summarize(mean_epa = mean(EPA),
            success_rate = mean(success),
            ypc=mean(yards_gained),
            plays=n()) %>%
  arrange(desc(mean_epa)) %>%
  filter(plays>50) %>%
  head(n=15L)
  

oregon_year[,3:5]<-round(oregon_year[,3:5],2)

oregon_year %>%
kable(caption = "Oregon Annual Efficient Running Seasons") %>%
  kable_styling(font_size = 10,bootstrap_options = c("condensed")) %>%
  footnote(general = "Top 15 Oregon rushing season or explosive runners")

```






```{r a5,echo=FALSE,message=FALSE,warning=FALSE}

ggfootball <- function(left_endzone_color = "gray90",
                       right_endzone_color = "gray90",
                       field_color = "green4",
                       field_alpha = 0.85,
                       top_buffer = 1,
                       bottom_buffer = 1,
                       left_buffer = 1,
                       right_buffer = 1,
                       five_yd_lines = TRUE,
                       ydline_labels = TRUE,
                       ydline_label_size = 4,
                       outer_hash = TRUE,
                       inner_hash = FALSE) {

  # Make middle of field green
  gplot <- ggplot2::ggplot() + ggplot2::geom_rect(data = NULL,
                    ggplot2::aes(xmin = 10, xmax = 110, ymin = 0, ymax = 53.3),
                    fill = field_color, color = "black", alpha = field_alpha) +

    # Add endzones
    ggplot2::geom_rect(data = NULL,
              ggplot2::aes(xmin = 0, xmax = 10, ymin = 0, ymax = 53.3),
              fill = left_endzone_color, color = "black") +
    ggplot2::geom_rect(data = NULL,
              ggplot2::aes(xmin = 110, xmax = 120, ymin = 0, ymax = 53.3),
              fill = right_endzone_color, color = "black") +

    # Format gridlines, tick marks, tick labels, and border of plot window
    ggplot2::theme_bw() +
    ggplot2::theme(panel.grid.minor = ggplot2::element_blank(),
          panel.grid.major = ggplot2::element_blank(),
          panel.border = ggplot2::element_blank(),
          axis.text = ggplot2::element_blank(),
          axis.ticks = ggplot2::element_blank(),
          axis.title = ggplot2::element_blank(),
          text = ggplot2::element_text(size = 16),
          #, legend.position = "none" # Optional hiding of legend
    ) +

    # Add x and y axis limits
    ggplot2::lims(x = c(0 - left_buffer, 120 + right_buffer),
                  y = c(0 - bottom_buffer, 53.3 + top_buffer))

  # Add vertical lines at each 5-yard increment
  if(five_yd_lines) {
    # Create data frame with necessary x and y coordinates
    five_yard_df <- data.frame(x = seq(from = 15, to = 105, by = 5))
    # Add to existing plot
    gplot <- gplot +
      ggplot2::geom_segment(data = five_yard_df,
                            mapping = ggplot2::aes(x = x, xend = x,
                                                   y = -Inf, yend = 53.3),
                            color = "white")
  }

  # Add yardline labels
  if(ydline_labels) {
    # Create data frame with labels and coordinates
    yard_labels_df <- data.frame(x = seq(from = 20, to = 100, by = 10),
                                 y = rep(x = 4, n = 9),
                                 digits = c(seq(from = 10, to = 50, by = 10),
                                            seq(from = 40, to = 10, by = -10)))
    # Add to existing plot
    gplot <- gplot +
      ggplot2::geom_text(data = yard_labels_df,
                         mapping = ggplot2::aes(x = x, y = y, label = digits),
                         color = "white", size = ydline_label_size)
    gplot <- gplot +
      ggplot2::geom_text(data = yard_labels_df,
                         mapping = ggplot2::aes(x = x, y = 53.3 - y,
                                                label = digits),
                         color = "white", angle = 180, size = ydline_label_size)
  }

  # Add outer hash marks to field
  if(outer_hash) {
    # Create data frame with hash mark x-coordinates
    hash_df <- data.frame(x = 11:109)
    # Add to existing plot
    gplot <- gplot +
      ggplot2::geom_segment(data = hash_df,
                            mapping = ggplot2::aes(x = x, xend = x,
                                                   y = 0.5, yend = 1.5),
                            color = "white") +
      ggplot2::geom_segment(data = hash_df,
                            mapping = ggplot2::aes(x = x, xend = x,
                                                   y = 51.8, yend = 52.8),
                            color = "white")
  }

  # Add inner hash marks to field
  if(inner_hash) {
    # Create data frame with hash mark x-coordinates
    hash_df <- data.frame(x = 11:109)
    # Add to existing plot
    gplot <- gplot +
      ggplot2::geom_segment(data = hash_df,
                            mapping = ggplot2::aes(x = x, xend = x,
                                                   y = 17.8, yend = 18.8),
                            color = "white") +
      ggplot2::geom_segment(data = hash_df,
                            mapping = ggplot2::aes(x = x, xend = x,
                                                   y = 34.6, yend = 35.6),
                            color = "white")
  }

  # Create final solid black outlines for the field
  gplot <- gplot +
    ggplot2::geom_rect(data = NULL,
                       ggplot2::aes(xmin = 10, xmax = 110,
                                   ymin = 0, ymax = 53.3),
                       fill = NA, color = "black") +
    ggplot2::geom_rect(data = NULL,
                       ggplot2::aes(xmin = 0, xmax = 120,
                                    ymin = 0, ymax = 53.3),
                       fill = NA, color = "black")

  # Return plot
  gplot
}

```

# Predicting Future Oregon Rushing Seasons
we will try to predict rushing attack for players on the roster next year. After some trial and error we found that mean_epa, ypc, games (career games played), yards_game would yield a 98% R-squared and a p-value less than 0.05, which is a sufficient bare bones model. 
```{r mod1,echo=FALSE,message=FALSE,warning=FALSE}


oregon_Pred<-o_rush_pbp %>%
  filter(pos_team == "Oregon",rush==1,down<=4,complete.cases(EPA)) %>%
  group_by(rusher_player_name,year) %>%
  summarize(mean_epa = mean(EPA),
            success_rate = mean(success),
            ypc=mean(yards_gained),
            plays=n(),
            total_success = sum(success),
            yards = sum(yards_gained),
            games= n_distinct(game_id),
            yards_gm = sum(yards_gained)/n_distinct(game_id)) %>%
  arrange(desc(rusher_player_name,year)) %>%
  filter(plays>30) 

oregon_Pred<-oregon_Pred %>%
  mutate(lag_mean_epa = ifelse(rusher_player_name == lag(rusher_player_name), lag(mean_epa), NA),
         lag_success_rate = ifelse(rusher_player_name == lag(rusher_player_name), lag(success_rate), NA),
         lag_ypc = ifelse(rusher_player_name == lag(rusher_player_name), lag(ypc), NA),
         lag_yards = ifelse(rusher_player_name == lag(rusher_player_name), lag(yards), NA))

oregon_Pred<-oregon_Pred %>%
  select(yards,mean_epa,success_rate,ypc,plays,total_success,games,yards_gm,lag_mean_epa,lag_success_rate,lag_ypc,lag_yards)

oregon_Pred<-oregon_Pred[,2:13]

#o_mod1<-lm(yards ~ .,data = oregon_Pred)
#summary(o_mod1)

### remove ##
o_mod2<-lm(yards ~ mean_epa + ypc + games + yards_gm,data = oregon_Pred)
summary(o_mod2)



```
#### Predicting 2021 Oregon Rushing Attack
We predict Tyler Shough to rush close to 400 yards, which we think he can easily hit given he rushed for 400 yards in a full season. CJ Verdell at nearly 1,800 yards would be nearly 700 more than his 2019 season. We don't think that will having even with the other two scenarios reducing their mean EPA (80% of their career average) and Yards per game (70% of career average). Sean Dollars showed a lot of promise and he will most certainly eat carries from Travis Dye (~1,400). Travis Dye is very high given he solely rushed in 2020 given CJ's injury. We'd be estatic if they both rushed over 1,000 yards but the unknown variable of Sean Dollars and if Anthony Brown will start may eat rush yards from all of them. 
```{r pred1,echo=FALSE,message=FALSE,warning=FALSE}

#o<-subset(o_rush_pbp,o_rush_pbp$pos_team=="Oregon")
#o<-subset(o,o$year=="2020")
#unique(o$rusher_player_name)

#situation 1
oregon_pred1<-o_rush_pbp %>%
  filter(pos_team == "Oregon",rush==1,down<=4,complete.cases(EPA)) %>%
  group_by(rusher_player_name) %>%
  summarize(mean_epa = mean(EPA),
            success_rate = mean(success),
            ypc=mean(yards_gained),
            games= n_distinct(game_id),
            yards_gm = sum(yards_gained)/n_distinct(game_id)) %>%
  arrange(desc(rusher_player_name,year)) %>%
  filter(rusher_player_name %in% c('CJ Verdell','Tyler Shough','Travis Dye')) 

#80% of mean EPA
oregon_pred2<-o_rush_pbp %>%
  filter(pos_team == "Oregon",rush==1,down<=4,complete.cases(EPA)) %>%
  group_by(rusher_player_name) %>%
  summarize(mean_epa = mean(EPA)*.8,
            success_rate = mean(success),
            ypc=mean(yards_gained),
            games= n_distinct(game_id),
            yards_gm = sum(yards_gained)/n_distinct(game_id)) %>%
  arrange(desc(rusher_player_name,year)) %>%
  filter(rusher_player_name %in% c('CJ Verdell','Tyler Shough','Travis Dye')) 

#70% of mean yards per game
oregon_pred3<-o_rush_pbp %>%
  filter(pos_team == "Oregon",rush==1,down<=4,complete.cases(EPA)) %>%
  group_by(rusher_player_name) %>%
  summarize(mean_epa = mean(EPA)*.8,
            success_rate = mean(success),
            ypc=mean(yards_gained),
            games= n_distinct(game_id),
            yards_gm = (sum(yards_gained)/n_distinct(game_id))*.7) %>%
  arrange(desc(rusher_player_name,year)) %>%
  filter(rusher_player_name %in% c('CJ Verdell','Tyler Shough','Travis Dye')) 

#avg. 
oregon_pred1$o_avg_pred<-predict.lm(o_mod2,oregon_pred1)
o_avg_pred_epa_80pct<-predict.lm(o_mod2,oregon_pred2)
o_avg_pred_ydgm_70pct<-predict.lm(o_mod2,oregon_pred3)

oregon_pred1$o_avg_pred_epa_80pct<-o_avg_pred_epa_80pct
oregon_pred1$o_avg_pred_ydgm_70pct<-o_avg_pred_ydgm_70pct

#oregon_pred1<-oregon_pred1 %>% select(-o_avg_pred_ypc_80pct)

oregon_pred4<-oregon_pred1 %>%
  select(rusher_player_name,games,o_avg_pred,o_avg_pred_epa_80pct,o_avg_pred_ydgm_70pct)

oregon_pred4[,3:5]<-round(oregon_pred4[,3:5],1)
names(oregon_pred4)[1]<-"Rushers"
names(oregon_pred4)[3]<-"Predicted Rushing Yards"
names(oregon_pred4)[4]<-"Predicted Rushing Yards @ 80% EPA"
names(oregon_pred4)[5]<-"Predicted Rushing Yards @ 70% Yards/GM"

oregon_pred4 %>%
kable(caption = "2021 Oregon Predicted Rushing Yards") %>%
  kable_styling(font_size = 10,bootstrap_options = c("condensed")) %>%
  footnote(general = "Missing Anthony Brown & Sean Dollars; not sufficient enough samples size. Predicted Rushing Yards @ 80% EPA takes 80% of runners expected play average, Predicted Rushing Yards @ 70% Yards/GM takes 70% of runners yards per game to account for Anthony Brown and Sean Dollars rushing")


```




